---
title: "assignment 5"
output: 
    github_document
editor_options: 
  chunk_output_type: inline
---

# Homework 05: Factor and figure management
## Part 1: Factor Management
With the data set of your choice, after ensuring the variable(s) you’re exploring are indeed factors, you are expected to:

Drop factor / levels;
Reorder levels based on knowledge from data.

```{r}
# load required libraries
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(scales))
```

### Drop Oceania
Filter the Gapminder data to remove observations associated with the continent of Oceania. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and Oceania; address the number of rows and the levels of the affected factors.

To drop the continent of Oceania, apply `filter()`
```{r}
gapminder_no_oceania <- gapminder %>% 
  filter(continent != "Oceania")
```

To take a glance at gapminder_no_oceania:
```{r}
str(gapminder_no_oceania)
```

To show the continents in gapminder_no_oceania:
```{r}
unique(gapminder_no_oceania$continent)
```
We can see from the result that Oceania is removed from gapminder. However, levels remain the same.

To remove the unused factor levels:
```{r}
gapminder_no_oceania <- gapminder_no_oceania %>% 
  # drop Oceania level since it's unused
  droplevels()

unique(gapminder_no_oceania$continent)
```
Now Oceania level is removed from levels.

```{r}
str(gapminder_no_oceania)
```

Before `droplevels()`, gapminder_no_oceania data frame has 142 levels for country, 5 levels for continent. Since continent Oceania has two countries, after `droplevels()`, the country level decreased to 140 and continent level decreased to 4. 

### Reorder the levels of continent

I tried to reorder the continent with standard deviation of lifeExp over years. Before reordering, let's take a look at the levels of continent first. 
```{r}
levels(gapminder_no_oceania$continent) %>% 
  head()
```
```{r}
# compute standard deviation for continents for later labelling
summ <- gapminder_no_oceania %>% 
  group_by(continent) %>% 
  summarise(sd = sd(lifeExp))
knitr::kable(summ)
```

Before apply factor re-leveling, first try to see the effect for `arrange()`.
```{r Part 1-1, fig.cap = "arrange() effect on plot before re-leveling"}
gapminder_no_oceania %>% 
  arrange(desc(lifeExp)) %>% 
  ggplot(aes(continent, lifeExp))+
  geom_point(aes(color = continent), alpha = 0.1) +
  geom_boxplot(aes(fill = continent), alpha = 0.25) +
  # label the standard deviation since it cannot be seen from boxplot clearly
  geom_label(data = summ, aes(x = continent, y = sd, 
    label = paste("SD: ", round(sd, 1)))) + 
  ggtitle("Effect of arrange() before releveling") + 
  theme_bw()
```

From the plot, standard deviation can somehow related to the box size. However, to be clear, I use `geom_lable()` to label the standard deviation values. We can see that `arrange()` doesn't have any effect on the order on the plot.

Apply `fct_reorder`, the result is:
```{r}
reordered_level <- fct_reorder(
  gapminder_no_oceania$continent,
  gapminder_no_oceania$lifeExp,
  sd) %>%
  levels() %>% 
  head()
reordered_level
```

To prove the result, `summarise()` is used:
```{r}
check_level <- gapminder_no_oceania %>% 
  group_by(continent) %>% 
  summarise(sd_lifeExp = sd(lifeExp)) %>% 
  arrange(sd_lifeExp) %>% 
  head()
knitr::kable(check_level)
```

If we compare check_level and reordered_level side by side, we can determine if `fct_reorder()` is correct or not:
```{r Part 1-3, fig.cap = "side by side comparison of level change"}
textplot(cbind(check_level, reordered_level))
```
From the result, the arranged country list is exactly same as reordered country level. So we can say `fct_reorder()` is giving correct result.

Use `fct_reorder()` for a plot:
```{r Part 1-2, fig.cap = "fct_reorder() effect on plot"}
# Change level order based on standard deviation of lifeExp
gapminder_no_oceania <- gapminder_no_oceania %>% 
  mutate(continent = fct_reorder(continent, lifeExp, sd))
gapminder_no_oceania %>% 
  ggplot(aes(continent, lifeExp)) +
  geom_point(aes(color = continent), alpha = 0.1) +
  geom_boxplot(aes(fill = continent), alpha = 0.25) + 
  geom_label(data = summ, aes(x = continent, y = sd, 
    label = paste("SD: ", round(sd, 1)))) + 
  ggtitle("Effect of fct_reorder() on plot") + 
  theme_bw()
```
From the plot, we can see that the continent now is ranked by the max lifeExp(top point of each continent). So `fct_reorder()` does change the continent order of the plot.


## Part 2: File I/O
### write_csv()

Before writing, check the levels of continents in gapminder_no_oceania and it is no longer in alphabetical order.
```{r}
# Reorder data frame by continent so it's not alphabetical based
gapminder_reordered <- gapminder %>%
  mutate(continent = fct_reorder(continent, lifeExp, sd)) %>%
  arrange(continent)
levels(gapminder_reordered$continent)
```

To explore `write_csv()`, we can write gapminder_no_oceania from part 1 into csv file
```{r}
write_csv(gapminder_reordered, "reordered.csv")
```

### read_csv()

To explore `read_csv()`, we can read data from no_oceania.csv
```{r}
data_read <- read_csv("reordered.csv")
knitr::kable(head(data_read, 10))
# Check if data matched
if (data_read == gapminder_reordered) print("Data read and written matched")
# overview of data read
glimpse(data_read)
```
From `glimpse()`, it can be seen that the first continent is Oceania and the continent order is same as original  data frame which is not alphabetic based. For the levels, the factors type has been changed into chr, so `levels()` will return NULL instead.


### saveRDS()
```{r}
saveRDS(gapminder_reordered, "reordered.rds")
```

### readRDS()
```{r}
read_rds = readRDS("reordered.rds")
if (read_rds == gapminder_reordered) print("Read RDS from file is same as the original one")
glimpse(read_rds)
knitr::kable(head(read_rds, 10))
```
From the result, we can see that `readRDS()` did not change the order of original reordered data frame.

```{r}
levels(read_rds$continent)
```
The level of continent factors remains the same.

### dput()
```{r}
dput(gapminder_reordered, "reordered.txt")
```

### dget()
```{r}
read_txt <- dget("reordered.txt")

if (read_txt == gapminder_reordered) print("Read txt from file is same as the original one")
# overview of the data read
glimpse(read_txt)
knitr::kable(head(read_txt, 10))
```

```{r}
levels(read_txt$continent)
```
From the result, level of continent factors also remain the same. 


## Part 3: Visualization design

I would like to show a plot of gdpPercap against population for selected four countries in two continents. Before visualization, the data can be shown and the trend of gdpPercap and population is obvious to observe. 
```{r Part 3-1, fig.cap = "Before visualization design"}
gapminder %>% 
  # select some countries to study
  filter(country == "Germany" | country == "Spain" | country == "Canada" | country == "United States") %>% 
  ggplot(aes(pop, gdpPercap)) + 
  geom_point(aes(color = country)) + 
  ggtitle("gdpPercap vs pop for four countries before visualization design")
```

After visualization design, we can give more information and better visualization on the plot.
```{r Part 3-2, figure.cap = "After visualization design"}
visualized_plot <- gapminder %>% 
  filter(country == "Germany" | country == "Spain" | country == "Canada" | country == "United States") %>% 
  ggplot(aes(pop, gdpPercap)) + 
  geom_point(aes(size = lifeExp, shape = country, color = country), alpha = 0.5) + 
  # scale since x has large data
  scale_x_log10(labels = comma_format()) + 
  facet_wrap(~continent) + 
  xlab("Population") + 
  ylab("Gdp Per Capita") + 
  theme(strip.background = element_rect(fill = "green"),
        axis.text = element_text(size = 8)) + 
  ggtitle("gdpPercap vs pop for four countries before visualization design")

visualized_plot
```
The resulting graph not only provides better visualization, but also gives extra info regarding life expectancy. It can be seen that life expectancy also has positive relationship with both gdpPercap and population. 

### plotly()
```{r}
# comment out for generating github_document, since ggplotly generates html output
#ggplot_visualized <- ggplotly(visualized_plot)
#ggplot_visualized
# Publish the graphs to Plotly with api_create()
#api_create(ggplot_visualized, filename="hw05-visualized-ggplotly")
```
[uploaded plotly](https://plot.ly/~yihaoz/1/#/)

`ggplotly()` converts ggplot2 charts to interactive d3 charts with good parity. Plotly also has WebGL and 3D data visualization support. For my plot, it gives more interaction. When the cursor is moved onto the points, it will show the details (country, lifeExp, pop, gdpPercap) of the that point. 

## Part 4: Writing figures to file
```{r}
ggsave("visualized_plot.png", visualized_plot, width=20, height=15, units = "cm", dpi = 500)
```

Here is the saved data:


## But I want to do more!
Make a deeper exploration of the forcats packages.

`fct_infreq()` will reorder factors levels by first apprearane or frequency.
```{r}
ordered <- gapminder$continent %>% 
  fct_infreq() %>%
  levels()
ordered
```

`fct_rev()` will reverse order of factor levels
```{r}
reversed <- gapminder$continent %>% 
  fct_infreq() %>%
  # to reverse the original order
  fct_rev() %>% 
  levels()
reversed
```

Clear side by side comparison:

```{r Part 5-3, fig.cap = "side by side comparison of reversed order"}
# use textplot for side by side print
textplot(cbind(ordered, reversed))
```


Use fct_reorder2() when you have a line chart of a quantitative x against another quantitative y and your factor provides the color. This way the legend appears in some order as the data! Contrast the legend on the left with the one on the right. [Reference](http://stat545.com/block029_factors.html#change-order-of-the-levels-principled)

```{r Part 5, figure.cap = "Population vs year before reorder"}
selected_countries <- c("Japan", "Canada", "Spain", "Germany")
gap <- gapminder %>%
  filter(country %in% selected_countries) %>% 
  droplevels()
ggplot(gap, aes(x = year, y = pop, color = country)) +
  geom_line() + 
  ggtitle("Population vs year before reorder") + 
  theme_bw()

# improved plot with fct_reorder2()
ggplot(gap, aes(x = year, 
                y = pop,
                color = fct_reorder2(country, year, pop))) +
  geom_line() +
  labs(color = "country") + 
  ggtitle("Population vs year after reorder") + 
  theme_bw()
```
The second plot legends are ranked by population now.
